### 2016-09-15 Chapters/ContinuumSolvationModels.tex
Finishing up response theory chapter
Almost done
Save the day's work
Polish

Continuum models have a venerable history in quantum chemistry:
\citeauthor{Onsager1936-wf}'s model\autocite{Onsager1936-wf} appeared in
1936 and the first form of the \acrshort{PCM} entered the stage in
1981.~\autocite{Miertus1981-mm}
Section \ref{sec:CSM-why-how} will offer a heuristic "derivation" of
continuum models.
While it cannot be thought as formally rigorous, it highlights the major
physical insights that have informed the creation of continuum models.
Still nowadays much research activity is expended on the
\acrshort{PCM}.~\autocite{Tomasi2011-un, Mennucci2012-dv, Lipparini2016-mo}
I will point out throughout the Chapter to developments outside our
group that are particularly interesting.

@@

Following \citeauthor{Tomasi2007-es}, we introduce the complete system
Hamiltonian in the \acrshort{BO} approximation:~\autocite{Tomasi2004-dc,
Tomasi2007-es}
\begin{equation}\label{eq:solution-ham}
 H(\vect{r}_\mathrm{S}, \vect{r}_\mathrm{E}) =
  H_\mathrm{S}(\vect{r}_\mathrm{S}) +  H_\mathrm{E}(\vect{r}_\mathrm{E})
+ H_\mathrm{SE}(\vect{r}_\mathrm{S}, \vect{r}_\mathrm{E})
\end{equation}
The Hamiltonian features solute terms, marked by the S subscript,
environment terms, marked by the E subscript, and interaction terms.
The coordinates $(\vect{r}_\mathrm{S}, \vect{r}_\mathrm{E})$ refer to
\emph{both} nuclei and electrons.
The interaction term is given by the usual Coulomb electrostatic
Hamiltonian.

One can replace the pure environment and interaction terms with their
classical counterparts and obtain the quantum/classical, possibly
\emph{polarizable}, Hamiltonian for an explicit \acrshort{QM}/\acrshort{MM} model.
This brings about the first important point: the need for statistical
averaging.
Whenever a large number of degrees of freedom is involved, one can
access macroscopic observables of the system by the appropriate
averaging of the microscopically detailed motion over phase space
trajectories or on the appropriate statistical
ensemble.~\autocite{Hill1960-ql}
The need for ensemble averages leads us to the following observations:
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
 \item
   the need for \emph{macroscopic} parameters, absent from the
   microscopic Hamiltonian Eq.~\eqref{eq:solution-ham}, in carrying out
   statistical simulations.
 \item
   Chosen a thermodynamic ensemble, the basic energetic quantity is
   accordingly determined. For example, the Gibbs free energy  $G$ is
   intrinsically related to the $(NpT)$ ensemble.
 \item
   Even atomistic simulations will lead to results that are essentially
   averages. The discrete picture of the system has been replaced by
   some continuous distribution function.
\end{enumerate*}

These observations strongly suggest that one can perform the averaging
step \emph{before} embarking into the solution of the quantum mechanical
problem and replace the full Hamiltonian with and \emph{effective}
Hamiltonian in which the environment degrees of freedom are properly
averaged.~\autocite{Angyan1992-vo, Tapia1992-pu}
The solute-environment interaction term would then be replaced by
a term only depending on the solute degrees of freedom, represented by
the averaged, continuous response functions of the solvent.
This is the physical setting of the \gls{PCM} and is schematically
depicted in Figure \ref{fig:IEF}.
The solute molecule is enclose in a \emph{cavity} inside the
structureless continuum representing the solvent and the basic energetic
quantity is a \emph{free energy}:\autocite{Tomasi1994-gv,
Amovilli1998-pv, Tomasi2005-vm}
\begin{equation}\label{eq:free-energy}
 G = G_\mathrm{cav} + G_\mathrm{el} + G_\mathrm{dis} + G_\mathrm{rep} + G_\mathrm{Mm}(T).
\end{equation}
The decomposition of the solute free energy takes into account the
different physical origins of the different solute-solvent interactions.
$G_\mathrm{cav}$ is the cavity formation free energy, a constant as long
as the solute geometry does not change.\autocite{Tomasi1994-gv,
Floris1997-tu}
The second term $G_\mathrm{el}$ is the most important one in the
framework of continuum models. It accounts for the mutual electrostatic
polarization of solute and solvent, see Section \ref{sec:IEF}.
The third $G_\mathrm{dis}$ and fourth $G_\mathrm{rep}$ terms account for
dispersion and repulsion interactions,
respectively.\autocite{Amovilli1997-fo, Weijo2010-jy}
Finally, $G_\mathrm{Mm}(T)$ accounts for the collective molecular
motions of the solute-solvent system. It explicitly depends on the
temperature $T$ and can be calculated from the solution partition
function, see \noparcite[ref.][]{Amovilli1998-pv} for further details.

@@

The molecular solute, represented by its charge density
$\rhoi$ is enclosed in a \emph{cavity} $\Omegai$.
The boundary of the cavity $\Gamma\equiv\partial\Omegai$ is a
2-manifold and is assumed to be continuously differentiable, \ie $\mathcal{C}^1$.
The cavity is carved out of an infinite, structureless, continuum
characterized by the bulk properties of the solvent.
The continuum fully spans the external subdomain $\Omegae$, see Fig. \ref{fig:IEF}.
The source charge density is not assumed to be continuous: both
classical point multipoles and quantum mechanical charge densities can
be treated within the model.
However, we assume that its support is entirely within the cavity,
$\Supp(\rhoi) \subseteq \Omegai$. This assumption is, of course, false
for quantum mechanical charge densities. It can however be proven that
the errors induced by this \emph{charge penetration} are not so
severe.~\autocite{Chipman2000-us, Cances2001-qs, Cances2001-qn}

@@

Given the partition of Euclidean space $\mathbb{R}^3$ outlined above,
find the solution $u(\vect{r})$ to the following set:
\begin{subequations}\label{eq:transmission}
  \begin{align}[left={\empheqlbrace}]
  \Li u &= \nabla^2 u = -4\pi\rhoi \,\, \text{in}\,\, \Omegai \label{eq:internal} \\
  \Le u &= 0 \,\, \text{in}\,\, \Omegae \label{eq:external} \\
  [u] &= \ue - \ui = 0 \,\, \text{on}\,\, \Gamma
  \label{eq:trace-jump} \\
[\partial_L u] &= \partiale u - \partiali u = 0 \,\, \text{on}\,\, \Gamma \label{eq:conormal-jump} \\
|u(\vect{r})| &\leq C \|\vect{r} \|^{-1} \,\,\text{for}\,\,\| \vect{r} \|\rightarrow\infty
\label{eq:radiation}
\end{align}
\end{subequations}
Eqs.~\eqref{eq:trace-jump} and \eqref{eq:conormal-jump}
are the \emph{jump conditions}, expressed in terms of trace operators
for the solution $u$ and its \emph{conormal} derivative.
For notational simplicity, will use the symbols $\partiale$ and $\partiali$
for the latter and only give it in explicit form when needed.
Further mathematical details and technical results on the definition of
function traces and their use in setting up the proper normal and
conormal derivatives can be found in the excellent book by
\citeauthor{Sauter2011-an}.~\autocite{Sauter2011-an}.
The fundamental solutions, or Green's functions, for the elliptic
differential operators $\Li$ and $\Le$ will be denoted by $\Gi$ and
$\Ge$, respectively.
As such the problem is formulated in the so-called \emph{strong} form:
the sought-after solution $u$ has to be at least two times continuously
differentiable.
Moreover, the solution is a function supported over $\mathbb{R}^3$ which
poses challenges to the numerical solution of the problem.
We will take a different approach and reformulate it in terms of
\emph{boundary integral operators}, \ie transform the \acsp{PDE} into a
\gls{BIE}.

The first step is to introduce the relevant boundary integral operators for the
above mentioned transmission problem.
They are three of the four components of the Calder\'on projector.
For any function $v$ in the suitable function space and for $\vect{s}, \vect{s}^\prime \in \Gamma$:
\begin{subequations}
\begin{align}
  (\bi{S}_\star v)(\vect{s}) &= \int_\Gamma
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff
  \vect{s}^\prime \label{eq:S} \\
  (\bi{D}_\star v)(\vect{s}) &= \int_\Gamma \partial_{L_\star,\vect{s}}
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff
  \vect{s}^\prime \label{eq:D} \\
  (\bi{D}^\dagger_\star v)(\vect{s}) &= \int_\Gamma
\partial_{L_\star,\vect{s}^\prime}
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff
  \vect{s}^\prime \label{eq:D-dagger}
\end{align}
\end{subequations}
where $\star$ is a placeholder for the $\mathrm{i}$ or $\mathrm{e}$ subscript.
Our next step is to rewrite the $u$ as the sum of two contributions:
\begin{equation}
  u(\vect{r}) = \esp(\vect{r}) + \xi(\vect{r}) = \int_{\Omegai}
  \frac{\rhoi(\vect{r}^\prime)}{|\vect{r}-\vect{r}^\prime|}
  + \xi(\vect{r})
\end{equation}
the former is the electrostatic potential of $\rhoi$ \emph{in vacuo}
(Newton potential), while the latter is the \emph{reaction} potential.
The reaction potential describes the polarization in the medium.
Moreover, it admits an integral representation as a single-layer
potential:
\begin{equation}
  \xi_\mathrm{i} = \bi{S}_\mathrm{i}\sigma
\end{equation}
where $\sigma$ is the \gls{ASC}, the key quantity in the \acrshort{PCM}.
As shown by \citeauthor{Cances1998-og}, the \acrshort{ASC} can be computed
as the \emph{unique} solution to the following integral equation:
\begin{equation}\label{eq:full-IEF}
  \left[ \bi{S}_\mathrm{e}\left(2\pi + \bi{D}^\dagger_\mathrm{i}\right)
  +
  \left(2\pi - \bi{D}_\mathrm{e}\right)\bi{S}_\mathrm{i}
  \right]\sigma =
  -\left[\left(2\pi-\bi{D}_\mathrm{e}\right)
  -\bi{S}_\mathrm{e}\bi{S}_\mathrm{i}^{-1}
  \left(2\pi-\bi{D}_\mathrm{i}\right)
  \right]\esp
\end{equation}
commonly called the \gls{IEF} equation.
The Green's functions for the elliptic differential operators
assume a central role in the boundary integral formulation and make the
\acrshort{IEF} equation valid for a wide range of external
environments: homogeneous isotropic, ionic and anisotropic
liquids,~\autocite{Cances1998-og}
and systems with interfaces.~\autocite{Corni2002-dr, Frediani2004-er,
Delgado2013-kd, DiRemigio2016-nn}

In the special and very common case of homogeneous and isotropic
environments, \ie regular dielectric materials characterized by a scalar
permittivity $\diel$, the Green's functions are:
\begin{alignat}{2}
  \Ge = \frac{1}{\diel|\vect{r}-\vect{r}^\prime|} = \frac{1}{\diel} \Gi,
  \quad&
  \Gi = \frac{1}{|\vect{r}-\vect{r}^\prime|}
\end{alignat}
yielding the isotropic \acrshort{IEF} equation:
\begin{equation}\label{eq:isotropic-IEF}
  \bi{R}_\diel\bi{S}\sigma = - \bi{R}_\infty\esp
\end{equation}
where the auxiliary operators:
\begin{alignat}{2}
  \bi{R}_\diel = \left[
  2\pi \left(\frac{\diel+1}{\diel-1}\right) - \bi{D}
  \right],
  \quad&
  \bi{R}_\infty =  \lim_{\diel\rightarrow\infty}\bi{R}_\diel
  = 2\pi-\bi{D}
\end{alignat}
have been introduced.
By letting the $\diel\rightarrow\infty$, one recovers the limiting case
where conductor boundary conditions are imposed. This leads to the
well-known \gls{COSMO} equation:~\autocite{Klamt1993-mj, Cossi2003-xe}
\begin{equation}
  \bi{S}\sigma = - f(\diel)\esp
\end{equation}
where the factor $f(\diel) = \frac{\diel-1}{\diel+x},\, 0\leq x \leq 1$
has been introduced to account for the error introduced by modelling the
dielectric as a perfect conductor.

@@

The molecular cavity is usually defined as a set of atom-centered
interlocking spheres. The radii are chosen from a suitable set, usually
van der Waals radii.~\autocite{Bondi1964-dt, Mantina2009-hb}
If one imagines to model a solvent molecule as a rigid sphere,
three types of molecular surfaces can be defined:
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
  \item
    the union of the interlocking spheres surfaces is the \gls{vdWS},
    Figure \ref{fig:vdWS}
  \item the locus of points defined by the center of the solvent
    spherical probe while rolling over the \acrshort{vdWS} is the
    \gls{SAS}, Figure \ref{fig:SAS}
  \item the locus of points defined by the contact point of the solvent
    spherical probe while rolling over the \acrshort{vdWS} is the \gls{SES}, Figure \ref{fig:SES}.
\end{enumerate*}
\begin{figure}[tb]
 \centering
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{vdWS.png}
   \caption{\acrlong*{vdWS} (\acrshort{vdWS})}
   \label{fig:vdWS}
 \end{subfigure}
 ~
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SAS.png}
   \caption{\acrlong*{SAS} (\acrshort{SAS})}
   \label{fig:SAS}
 \end{subfigure}

 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SES.png}
   \caption{\acrlong*{SES} (\acrshort{SES})}
   \label{fig:SES}
 \end{subfigure}
 \caption{Molecular surfaces from interlocking sets of atom-centered spheres.
 Reproduced from \noparcite[ref.][]{Harbrecht2011-dk}.}
\end{figure}
Alternative definitions based on the isodensity surface of the molecular solute
have been put forth.~\autocite{Foresman1996-wv}

Steps \ref{item:mesh} and \ref{item:discretize} are closely intertwined,
in the technical \acrshort{BEM} literature they go under the name of
\emph{finite element discretization}.
Let us define a mesh of the domain $\Gamma$ as an indexed collection of
intervals with non-zero measure $\lbrace I_i\rbrace_{i=1}^{N_\mathrm{mesh}}$, where
$N_\mathrm{mesh}$ is the size of the mesh.
Then, in a rather informal sense, a finite element is the
mathematical entity formed by the vector space of functions that are
piecewise regular on any given interval in the mesh.~\autocite{Ern2004-oo}
Basically, given an element in the mesh, we "attach" a basis set of
piecewise polynomials of a prescribed degree and use this basis to
provide a representation of quantities defined on $\Gamma$ within each
mesh interval.
Mesh generation for the molecular surfaces mentioned above is usually
achieved by modified triangulation procedures. In this thesis we used
the GePol algorithm~\autocite{Pascual-Ahuir1987-uo, Pascual-Ahuir1990-lp,
Pomelli1998-qp, Pomelli2001-sj, Frediani2004-ua, Pomelli2007-wq}
and the mesh generators of Harbrecht and Randrianarivony especially
tailored for the wavelet \acrshort{BEM}.~\autocite{Harbrecht2009-no,
Harbrecht2011-dk}

The finite element discretization is usually carried out either by the
\emph{collocation} or by the \emph{Galerkin} approach.
Indeed, it is possible to show that discretization by collocation
corresponds to a Galerkin method with a very specific (and asymmetric)
choice of \emph{trial} and \emph{test}
spaces.~\autocite{Hackbusch1995-uq, Ern2004-oo}
This step transforms the original integral equation into a system of
linear equations whose dimension relates to the underlying mesh size.
For general integral equations and finite element discretizations,
the \emph{stiffness matrix} obtained is dense and proper linear algebra
techniques need to be employed in order to avoid a detrimental impact on
computational performance.
In addition, finite element discretization might destroy intrinsic
symmetries of the original integral
operators. These can be recovered by appropriate \emph{a posteriori}
symmetrization procedures introducing, however, a degree of
arbitrariness in the \acrshort{BEM} procedure.~\autocite{Lange2011-eu}

>>>>

Continuum models have a venerable history in quantum chemistry:
\citeauthor{Onsager1936-wf}'s model\autocite{Onsager1936-wf} appeared in
1936 and the first form of the \acrshort{PCM} entered the stage in
1981.\autocite{Miertus1981-mm}
Section \ref{sec:CSM-why-how} will offer a heuristic "derivation" of
continuum models.
While it cannot be thought as formally rigorous, it highlights the major
physical insights that have informed the creation of continuum models.
Still nowadays much research activity is expended on the
\acrshort{PCM}.\autocite{Tomasi2011-un, Mennucci2012-dv, Lipparini2016-mo}
I will point out throughout the Chapter to developments outside our
group that are particularly interesting.

@@

Following \citeauthor{Tomasi2007-es}, we introduce the complete system
Hamiltonian in the \acrshort{BO} approximation:\autocite{Tomasi2004-dc,
Tomasi2007-es}
\begin{equation}\label{eq:solution-ham}
 H(\vect{r}_\mathrm{S}, \vect{r}_\mathrm{E}) =
  H_\mathrm{S}(\vect{r}_\mathrm{S}) +  H_\mathrm{E}(\vect{r}_\mathrm{E})
+ H_\mathrm{SE}(\vect{r}_\mathrm{S}, \vect{r}_\mathrm{E})
\end{equation}
The Hamiltonian features solute terms, marked by the S subscript,
environment terms, marked by the E subscript, and interaction terms.
The coordinates $(\vect{r}_\mathrm{S}, \vect{r}_\mathrm{E})$ refer to
\emph{both} nuclei and electrons.
The interaction term is given by the usual Coulomb electrostatic
Hamiltonian.

One can replace the pure environment and interaction terms with their
classical counterparts and obtain the quantum/classical, possibly
\emph{polarizable}, Hamiltonian for an explicit \acrshort{QM}/\acrshort{MM} model.
This brings about the first important point: the need for statistical
averaging.
Whenever a large number of degrees of freedom is involved, one can
access macroscopic observables of the system by the appropriate
averaging of the microscopically detailed motion over phase space
trajectories or on the appropriate statistical
ensemble.\autocite{Hill1960-ql}
The need for ensemble averages leads us to the following observations:
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
 \item
   the need for \emph{macroscopic} parameters, absent from the
   microscopic Hamiltonian Eq. \eqref{eq:solution-ham}, in carrying out
   statistical simulations.
 \item
   Chosen a thermodynamic ensemble, the basic energetic quantity is
   accordingly determined. For example, the Gibbs free energy  $G$ is
   intrinsically related to the $(NpT)$ ensemble.
 \item
   Even atomistic simulations will lead to results that are essentially
   averages. The discrete picture of the system has been replaced by
   some continuous distribution function.
\end{enumerate*}

These observations strongly suggest that one can perform the averaging
step \emph{before} embarking into the solution of the quantum mechanical
problem and replace the full Hamiltonian with and \emph{effective}
Hamiltonian in which the environment degrees of freedom are properly
averaged.\autocite{Angyan1992-vo, Tapia1992-pu}
The solute-environment interaction term would then be replaced by
a term only depending on the solute degrees of freedom, represented by
the averaged, continuous response functions of the solvent.
This is the physical setting of the \gls{PCM} and is schematically
depicted in Figure \ref{fig:IEF}.
The solute molecule is enclose in a \emph{cavity} inside the
structureless continuum representing the solvent and the basic energetic
quantity is a \emph{free energy}:\autocite{Tomasi1994-gv,
Amovilli1998-pv, Tomasi2005-vm}
\begin{equation}\label{eq:free-energy}
 G = G_\mathrm{cav} + G_\mathrm{el} + G_\mathrm{dis} + G_\mathrm{rep} + G_\mathrm{Mm}(T).
\end{equation}
The decomposition of the solute free energy takes into account the
different physical origins of the different solute-solvent interactions.
$G_\mathrm{cav}$ is the cavity formation free energy, a constant as long
as the solute geometry does not change.\autocite{Tomasi1994-gv,
Floris1997-tu}
The second term $G_\mathrm{el}$ is the most important one in the
framework of continuum models. It accounts for the mutual electrostatic
polarization of solute and solvent, see Section \ref{sec:IEF}.
The third $G_\mathrm{dis}$ and fourth $G_\mathrm{rep}$ terms account for
dispersion and repulsion interactions,
respectively.\autocite{Amovilli1997-fo, Weijo2010-jy}
Finally, $G_\mathrm{Mm}(T)$ accounts for the collective molecular
motions of the solute-solvent system. It explicitly depends on the
temperature $T$ and can be calculated from the solution partition
function, see \noparcite[ref.][]{Amovilli1998-pv} for further details.

@@

The molecular solute, represented by its charge density
$\rhoi$ is enclosed in a \emph{cavity} $\Omegai$.
The boundary of the cavity $\Gamma\equiv\partial\Omegai$ is a
2-manifold and is assumed to be continuously differentiable, \ie $\mathcal{C}^1$.
The cavity is carved out of an infinite, structureless, continuum
characterized by the bulk properties of the solvent.
The continuum fully spans the external subdomain $\Omegae$, see Fig. \ref{fig:IEF}.
The source charge density is not assumed to be continuous: both
classical point multipoles and quantum mechanical charge densities can
be treated within the model.
However, we assume that its support is entirely within the cavity,
$\Supp(\rhoi) \subseteq \Omegai$. This assumption is, of course, false
for quantum mechanical charge densities. It can however be proven that
the errors induced by this \emph{charge penetration} are not so
severe.\autocite{Chipman2000-us, Cances2001-qs, Cances2001-qn}

@@

Given the partition of Euclidean space $\mathbb{R}^3$ outlined above,
find the solution $u(\vect{r})$ to the following set:
\begin{subequations}\label{eq:transmission}
  \begin{align}[left={\empheqlbrace}]
  \Li u &= \nabla^2 u = -4\pi\rhoi \,\, \text{in}\,\, \Omegai \label{eq:internal} \\
  \Le u &= 0 \,\, \text{in}\,\, \Omegae \label{eq:external} \\
  [u] &= \ue - \ui = 0 \,\, \text{on}\,\, \Gamma
  \label{eq:trace-jump} \\
[\partial_L u] &= \partiale u - \partiali u = 0 \,\, \text{on}\,\, \Gamma \label{eq:conormal-jump} \\
|u(\vect{r})| &\leq C \|\vect{r} \|^{-1} \,\,\text{for}\,\,\| \vect{r} \|\rightarrow\infty
\label{eq:radiation}
\end{align}
\end{subequations}
Eqs. \eqref{eq:trace-jump} and \eqref{eq:conormal-jump}
are the \emph{jump conditions}, expressed in terms of trace operators
for the solution $u$ and its \emph{conormal} derivative.
For notational simplicity, will use the symbols $\partiale$ and $\partiali$
for the latter and only give it in explicit form when needed.
Further mathematical details and technical results on the definition of
function traces and their use in setting up the proper normal and
conormal derivatives can be found in the excellent book by
\citeauthor{Sauter2011-an}.\autocite{Sauter2011-an}.
The fundamental solutions, or Green's functions, for the elliptic
differential operators $\Li$ and $\Le$ will be denoted by $\Gi$ and
$\Ge$, respectively.
As such the problem is formulated in the so-called \emph{strong} form:
the sought-after solution $u$ has to be at least two times continuously
differentiable.
Moreover, the solution is a function supported over $\mathbb{R}^3$ which
poses challenges to the numerical solution of the problem.
We will take a different approach and reformulate it in terms of
\emph{boundary integral operators}, \ie transform the \acsp{PDE} into a
\gls{BIE}.

The first step is to introduce the relevant boundary integral operators for the
above mentioned transmission problem.
They are three of the four components of the Calder\'on projector.
For any function $v$ in the suitable function space and for $\vect{s}, \vect{s}^\prime \in \Gamma$:
\begin{subequations}
\begin{align}
  (\bi{S}_\star v)(\vect{s}) &= \int_\Gamma
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff
  \vect{s}^\prime \label{eq:S} \\
  (\bi{D}_\star v)(\vect{s}) &= \int_\Gamma \partial_{L_\star,\vect{s}}
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff
  \vect{s}^\prime \label{eq:D} \\
  (\bi{D}^\dagger_\star v)(\vect{s}) &= \int_\Gamma
\partial_{L_\star,\vect{s}^\prime}
  G_\star(\vect{s}, \vect{s}^\prime)v(\vect{s}^\prime)\diff
  \vect{s}^\prime \label{eq:D-dagger}
\end{align}
\end{subequations}
where $\star$ is a placeholder for the $\mathrm{i}$ or $\mathrm{e}$ subscript.
Our next step is to rewrite the $u$ as the sum of two contributions:
\begin{equation}
  u(\vect{r}) = \esp(\vect{r}) + \xi(\vect{r}) = \int_{\Omegai}
  \frac{\rhoi(\vect{r}^\prime)}{|\vect{r}-\vect{r}^\prime|}
  + \xi(\vect{r})
\end{equation}
the former is the electrostatic potential of $\rhoi$ \emph{in vacuo}
(Newton potential), while the latter is the \emph{reaction} potential.
The reaction potential describes the polarization in the medium.
Moreover, it admits an integral representation as a single-layer
potential:
\begin{equation}
  \xi_\mathrm{i} = \bi{S}_\mathrm{i}\sigma
\end{equation}
where $\sigma$ is the \gls{ASC}, the key quantity in the \acrshort{PCM}.
As shown by \citeauthor{Cances1998-og}, the \acrshort{ASC} can be computed
as the \emph{unique} solution to the following integral equation:
\begin{equation}\label{eq:full-IEF}
  \left[ \bi{S}_\mathrm{e}\left(2\pi + \bi{D}^\dagger_\mathrm{i}\right)
  +
  \left(2\pi - \bi{D}_\mathrm{e}\right)\bi{S}_\mathrm{i}
  \right]\sigma =
  -\left[\left(2\pi-\bi{D}_\mathrm{e}\right)
  -\bi{S}_\mathrm{e}\bi{S}_\mathrm{i}^{-1}
  \left(2\pi-\bi{D}_\mathrm{i}\right)
  \right]\esp
\end{equation}
commonly called the \gls{IEF} equation.
The Green's functions for the elliptic differential operators
assume a central role in the boundary integral formulation and make the
\acrshort{IEF} equation valid for a wide range of external
environments: homogeneous isotropic, ionic and anisotropic
liquids,\autocite{Cances1998-og}
and systems with interfaces.\autocite{Corni2002-dr, Frediani2004-er,
Delgado2013-kd, DiRemigio2016-nn}

In the special and very common case of homogeneous and isotropic
environments, \ie regular dielectric materials characterized by a scalar
permittivity $\diel$, the Green's functions are:
\begin{alignat}{2}
  \Ge = \frac{1}{\diel|\vect{r}-\vect{r}^\prime|} = \frac{1}{\diel} \Gi,
  \quad&
  \Gi = \frac{1}{|\vect{r}-\vect{r}^\prime|}
\end{alignat}
yielding the isotropic \acrshort{IEF} equation:
\begin{equation}\label{eq:isotropic-IEF}
  \bi{R}_\diel\bi{S}\sigma = - \bi{R}_\infty\esp
\end{equation}
where the auxiliary operators:
\begin{alignat}{2}
  \bi{R}_\diel = \left[
  2\pi \left(\frac{\diel+1}{\diel-1}\right) - \bi{D}
  \right],
  \quad&
  \bi{R}_\infty =  \lim_{\diel\rightarrow\infty}\bi{R}_\diel
  = 2\pi-\bi{D}
\end{alignat}
have been introduced.
By letting the $\diel\rightarrow\infty$, one recovers the limiting case
where conductor boundary conditions are imposed. This leads to the
well-known \gls{COSMO} equation:\autocite{Klamt1993-mj, Cossi2003-xe}
\begin{equation}
  \bi{S}\sigma = - f(\diel)\esp
\end{equation}
where the factor $f(\diel) = \frac{\diel-1}{\diel+x},\, 0\leq x \leq 1$
has been introduced to account for the error introduced by modelling the
dielectric as a perfect conductor.

@@

The molecular cavity is usually defined as a set of atom-centered
interlocking spheres. The radii are chosen from a suitable set, usually
van der Waals radii.\autocite{Bondi1964-dt, Mantina2009-hb}
If one imagines to model a solvent molecule as a rigid sphere,
three types of molecular surfaces can be defined:
\begin{enumerate*}[label={\alph*)},font={\color{PMS1797}}]
  \item
    the union of the interlocking spheres surfaces is the \gls{vdWS},
    Figure \ref{fig:vdWS}
  \item the locus of points defined by the center of the solvent
    spherical probe while rolling over the \acrshort{vdWS} is the
    \gls{SAS}, Figure \ref{fig:SAS}
  \item the locus of points defined by the contact point of the solvent
    spherical probe while rolling over the \acrshort{vdWS} is the \gls{SES}, Figure \ref{fig:SES}.
\end{enumerate*}
\begin{figure}[tb]
 \centering
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{vdWS.png}
   \caption{\acrlong*{vdWS} (\acrshort{vdWS})}
   \label{fig:vdWS}
 \end{subfigure}
 ~
 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SAS.png}
   \caption{\acrlong*{SAS} (\acrshort{SAS})}
   \label{fig:SAS}
 \end{subfigure}

 \begin{subfigure}{.45\textwidth}
   \centering
   \includegraphics[width=.85\textwidth]{SES.png}
   \caption{\acrlong*{SES} (\acrshort{SES})}
   \label{fig:SES}
 \end{subfigure}
 \caption{Molecular surfaces from interlocking sets of atom-centered spheres.
 Reproduced from \noparcite[ref.][]{Harbrecht2011-dk}.}
\end{figure}
Alternative definitions based on the isodensity surface of the molecular solute
have been put forth.\autocite{Foresman1996-wv}

Steps \ref{item:mesh} and \ref{item:discretize} are closely intertwined,
in the technical \acrshort{BEM} literature they go under the name of
\emph{finite element discretization}.
Let us define a mesh of the domain $\Gamma$ as an indexed collection of
intervals with non-zero measure $\lbrace I_i\rbrace_{i=1}^{N_\mathrm{mesh}}$, where
$N_\mathrm{mesh}$ is the size of the mesh.
Then, in a rather informal sense, a finite element is the
mathematical entity formed by the vector space of functions that are
piecewise regular on any given interval in the mesh.\autocite{Ern2004-oo}
Basically, given an element in the mesh, we "attach" a basis set of
piecewise polynomials of a prescribed degree and use this basis to
provide a representation of quantities defined on $\Gamma$ within each
mesh interval.
Mesh generation for the molecular surfaces mentioned above is usually
achieved by modified triangulation procedures. In this thesis we used
the GePol algorithm\autocite{Pascual-Ahuir1987-uo, Pascual-Ahuir1990-lp,
Pomelli1998-qp, Pomelli2001-sj, Frediani2004-ua, Pomelli2007-wq}
and the mesh generators of Harbrecht and Randrianarivony especially
tailored for the wavelet \acrshort{BEM}.\autocite{Harbrecht2009-no,
Harbrecht2011-dk}

The finite element discretization is usually carried out either by the
\emph{collocation} or by the \emph{Galerkin} approach.
Indeed, it is possible to show that discretization by collocation
corresponds to a Galerkin method with a very specific (and asymmetric)
choice of \emph{trial} and \emph{test}
spaces.\autocite{Hackbusch1995-uq, Ern2004-oo}
This step transforms the original integral equation into a system of
linear equations whose dimension relates to the underlying mesh size.
For general integral equations and finite element discretizations,
the \emph{stiffness matrix} obtained is dense and proper linear algebra
techniques need to be employed in order to avoid a detrimental impact on
computational performance.
In addition, finite element discretization might destroy intrinsic
symmetries of the original integral
operators. These can be recovered by appropriate \emph{a posteriori}
symmetrization procedures introducing, however, a degree of
arbitrariness in the \acrshort{BEM} procedure.\autocite{Lange2011-eu}

In the GePol procedure, the molecular surface is meshed by means of
$N_\mathrm{ts}$ spherical polygons $\Pi_i$ whose centroids define the
collocation points. One-point quadrature rules are then applied at the
centroid to obtain the matrix equation:
\begin{equation}
 \mat{R}^{-1}\mat{T} \vect{q}
  =
  \vect{v}.
\end{equation}
$\vect{q}$ and $\vect{v}$ are vectors of dimension $N_\mathrm{ts}$
collecting the values of the \acrshort{ASC} and \gls{MEP}
at the collocation points. $\mat{R}^{-1}\mat{T}$ is the stiffness
matrix.
The matrices representing the boundary integral operators are:
\begin{align}
  \mat{T} &=
  \left(2\pi\mat{I} - \mat{D}_\mathrm{e}\mat{A}\right)\mat{S}_\mathrm{i}
  +\mat{S}_\mathrm{e}\left(2\pi\mat{I} +
  \mat{A}\mat{D}_\mathrm{i}^\dagger\right) \\
  \mat{R} &=
  \left(2\pi\mat{I} - \mat{D}_\mathrm{e}\mat{A}\right) -
  \mat{S}_\mathrm{e}\mat{S}^{-1}_\mathrm{i}\left(2\pi\mat{I}-\mat{D}_\mathrm{i}\mat{A}\right)
\end{align}
$\mat{I}$ is the $N_\mathrm{ts}$-dimensional identity matrix while the
other matrices are defined as:
\begin{subequations}
  \begin{align}
    A_{ij} &= a_i\delta_{ij}, \quad a_i = \mathrm{area}(\Pi_i) \\
    S_{ij, \mathrm{i}} &=
    k\sqrt{\frac{4\pi}{a_i}}\delta_{ij} +
    \frac{1}{|\vect{s}_i-\vect{s}_j|}(1-\delta_{ij}) \\
    D_{ii, \mathrm{i}} &=
    -k\sqrt{\frac{\pi}{a_i}}\frac{1}{R_i}\delta_{ij}
    + \frac{(\vect{s}_i-\vect{s}_j)\cdot\vect{n}_j}{|\vect{s}_i-\vect{s}_j|^3}(1-\delta_{ij})
  \end{align}
\end{subequations}
in terms of the centroids $\vect{s}_i, \vect{s}_j$, the finite elements
areas and their curvatures $R_i$.
The factor $k=1.07$ is introduced to achieve a better precision on the
diagonal elements.\autocite{Tomasi2005-vm}
Expressions for $\mat{S}_\mathrm{e}$, $\mat{D}_\mathrm{e}$ and
$\mat{D}^\dagger_\mathrm{e}$ depend on the specific outer environment
and can be found in the literature.\autocite{Corni2002-dr, Frediani2004-er,
Tomasi2005-vm, Delgado2013-kd, DiRemigio2016-nn}
